---
title: "SHARE_Italy"
format:
  pdf:
    toc: true
    toc-depth: 3
    toc-title: Contents
    number-sections: true
    colorlinks: true
editor: visual
author: Sula Disem
---

```{r}
#| echo: false
#| message: false
here::i_am("Final_SHARE_Italy.Rproj")
library(here)
library(ggplot2)
library(tidyr)
library(dplyr)
library(vroom)
library(stringr)
library(knitr)
library(patchwork)
library(broom)
library(stats)
library(Gini)
library(ineq)
```

#### Link to the github repository of this project {.appendix}

[Github repository](https://github.com/suladi23/measurement_issues_SHARE_Italy)

#### Source of the data {.appendix}

This paper uses data from the generated easySHARE data set (DOI: [10.6103/SHARE.easy.800](https://share-eric.eu/data/data-set-details/easyshare)), see Gruber et al. (2014) for methodological details. The easySHARE release 8.0.0 is based on SHARE Waves 1, 2, 3 (SHARELIFE), 4, 5, 6, 7 and 8 (DOIs: [10.6103/SHARE.w1.800](http://dx.doi.org/10.6103/SHARE.w1.800), [10.6103/SHARE.w2.800](http://dx.doi.org/10.6103/SHARE.w2.800), [10.6103/SHARE.w3.800](http://dx.doi.org/10.6103/SHARE.w3.800), [10.6103/SHARE.w4.800](http://dx.doi.org/10.6103/SHARE.w4.800),[10.6103/SHARE.w5.800](http://dx.doi.org/10.6103/SHARE.w5.800), [10.6103/SHARE.w6.800](http://dx.doi.org/10.6103/SHARE.w6.800), [10.6103/SHARE.w7.800](http://dx.doi.org/10.6103/SHARE.w7.800), [10.6103/SHARE.w8.800](http://dx.doi.org/10.6103/SHARE.w8.800))

#### References {.appendix}

Börsch-Supan, A. & S. Gruber (2020): *easySHARE. Release version: 8.0.0. SHARE-ERIC. Dataset.* doi: [10.6103/SHARE.easy.800](http://dx.doi.org/10.6103/SHARE.easy.800) 
Istat 2018, *Condizione di vita, reddito e carico fiscale delle famiglie*
Asada, Y., Hurley, J., Norheim, O.F. et al. *A three-stage approach to measuring health inequalities and inequities*. Int J Equity Health 13, 98 (2014). https://doi.org/10.1186/s12939-014-0098-y

```{r}
#| echo: false
#| message: false
load("easySHARE_rel8_0_0.rda")
```

```{r}
#| echo: false
#| message: false
wave7 <- easySHARE_rel8_0_0 |>
  filter(wave == "7")
```

```{r}
#| echo: false
#| message: false
Italy <- wave7 |>
  filter(country == "16")
```

```{r}
#| echo: false
#| message: false
France <- wave7 |>
  filter(country == "17")
```

```{r}
#| echo: false
#| message: false
Germany <- wave7 |>
  filter(country == "12")
```

## Description of the Dataset for Italy

The SHARE survey, or Survey of Health, Ageing and Retirement in Europe, is a multidisciplinary and cross-national research initiative aimed at understanding the dynamics of health, aging, and retirement among older adults in Europe and beyond. It aims to provide insights into the health, social, economic, and environmental factors influencing individuals' well-being and quality of life as they age. To do so it collects data from the same individual over multiple waves at regular intervals, to track changes in various aspect of life over time. The data is collected from a wide range of European countries, to enable comparison with trends and patterns across different cultures and economic contexts.

SHARE employs a rigorous **sampling design** to ensure representativeness of the older population in each participating country. The survey includes individuals from diverse socio-economic backgrounds and geographic regions.

```{r warning=FALSE}
#| echo: false
#| message: false
nrows_Italy=nrow(Italy)
ncols_Italy=ncol(Italy)
```

After some manipulation of the data, filtering it to use only "Wave 7" and to consider only the country of choice "Italy", the Dataset we will be using has **`r nrows_Italy` rows** and **`r ncols_Italy` columns (variables)**

```{r}
#| echo: false
#| message: false
#Mainupulating the data to be able to work better afterwards

Italy$ep005_[Italy$ep005_ == 1] <- "Retired"
Italy$ep005_[Italy$ep005_ == 2] <- "Employed or Self-employed"
Italy$ep005_[Italy$ep005_ == 3] <- "Unemployed"
Italy$ep005_[Italy$ep005_ == 4] <- "Permanently Sick or Disabled"
Italy$ep005_[Italy$ep005_ == 5] <- "Homemaker"
Italy$ep005_[Italy$ep005_ == 97] <- "Other"

Italy$female[Italy$female == 0] <- "Male"
Italy$female[Italy$female == 1] <- "Female"

Italy <- Italy %>%
  rename(
    Gender = female,
    `Employment status` = ep005_
  )
```

The following table provides some descriptive statistics of Income, Age and Years of education.

```{r}
#| echo: false
#| message: false
# Calculate descriptive statistics for household income
Italy_income_filtered <- Italy %>% 
  filter(thinc_m != -10)
summary_income <- summary(Italy_income_filtered$thinc_m)

# Calculate descriptive statistics for gender
gender_stats <- table(Italy$Gender)

# Calculate descriptive statistics for age
summary_age <- summary(Italy$age)

# Calculate descriptive statistics for years of education
Italy_education_filtered <- Italy |>
  filter(eduyears_mod >= 0)
summary_education <- summary(Italy_education_filtered$eduyears_mod)

# Calculate descriptive statistics for employment status
Italy_employment_filtered <- Italy |>
  filter(`Employment status` != "-12" & `Employment status` != "-15")
employment_status <- table(Italy_employment_filtered$`Employment status`)
```

```{r}
#| echo: false
#| message: false
summary_combined <- data.frame(
  Variable = c("Income", "Age", "Years of education"),
  Min = c(summary_income["Min."], summary_age["Min."], summary_education["Min."]),
  `1st Qu.` = c(summary_income["1st Qu."], summary_age["1st Qu."], summary_education["1st Qu."]),
  Median = c(summary_income["Median"], summary_age["Median"], summary_education["Median"]),
  Mean = c(summary_income["Mean"], summary_age["Mean"], summary_education["Mean"]),
  `3rd Qu.` = c(summary_income["3rd Qu."], summary_age["3rd Qu."], summary_education["3rd Qu."]),
  Max = c(summary_income["Max."], summary_age["Max."], summary_education["Max."])
)
kable(summary_combined, format = "markdown")
```

Out of `r nrows_Italy` observations, we have here a table showing how the individuals are distributed among different types of employment status and gender:

```{r}
#| echo: false
#| message: false
combined <- data.frame(
  Count = c(employment_status, gender_stats))
kable(combined, format = "markdown")
```

## Measures of inequality and poverty based on household income

Let's now look into the average household income of Italy for wave 7 and compare it with two other countries: France and Germany.

```{r}
#| echo: false
#| message: false
country_names <- c("Italy", "France", "Germany")

# Calculate average household income for Italy
Italy_avg_income <- Italy %>%
  filter(thinc_m != -10) %>%
  summarise(Average_Income = mean(thinc_m))

# Calculate average household income for two other countries (France and Netherlands)
France_avg_income <- France %>%
  filter(thinc_m != 0) %>%
  summarise(Average_Income = mean(thinc_m))

Germany_avg_income <- Germany %>%
  filter(thinc_m != -10) %>%
  summarise(Average_Income = mean(thinc_m))

# Combine the results into a single data frame
avg_income_comparison <- rbind(Italy_avg_income, France_avg_income, Germany_avg_income)

avg_income_comparison <- cbind(Country = country_names, avg_income_comparison)

# Graph the average household income for the three countries
ggplot(avg_income_comparison, aes(x = Country, y = Average_Income, fill = Country)) +
  geom_bar(stat = "identity", width = 0.5) +
  labs(x = "Country", y = "Average Household Income", title = "Average Household Income Comparison") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

The average household income for Wave 7 in Italy, France, and Germany are graphically compared above.

Based on the comparison, **France** has the highest average income (`r France_avg_income$Average_Income`), followed not so far by **Germany** (`r Germany_avg_income$Average_Income`) and way lower by **Italy** (`r Italy_avg_income$Average_Income`)

In the SHARE release guide for wave 7, some issues in terms of the imputations of wave 7 for household income have been pointed out. To compute the total household income for household that had mixed interviews, respondedents to the other interview of "SHARELIFE" were treated as "nonresponding partners" (NRPs) in the regular interview. The problem is that detailed questions on current income were asked only to those with regular interview, so like standard NRPs issues, "incomes of all household members who answered the SHARELIFE interview are unobservable and focusing attention on the observable incomes of the regular respondents may introduce a severe downward bias in the aggregated measure of total household income.

In general, measuring household income is always difficult, because of misreporting or non response as cited before.

These issues could introduce bias and inaccuracies in household income estimates, leading to potentially unreliable rankings.

## Measuring income inequality in Italy

Let's first have a look into the distribution of income in Italy with the help of a graph to have a general view of it, after that we will get into the measuring of inequality in net income of households in Italy, using Gini, Theil and Palma ratio.

```{r}
#| echo: false
#| message: false
# Plotting a density plot of net income
ggplot(data = Italy_income_filtered, aes(x = thinc_m)) +
  geom_histogram(binwidth = 1000, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Net Income per households in Italy",
       x = "Net Income (thousands)",
       y = "Frequency") +
  scale_x_continuous(labels = scales::comma_format(scale = 1e-3)) +
  theme_minimal()
```

### Gini Coefficient

To measure inequality let's start by using the **Gini** coefficient, an index to measure how much the distribution of income between individuals deviates from a perfectly equal distribution (Worldbank, Metadata Glossary). It is an index, so it provides a result between 0 and 1, a higher Gini coefficient indicates higher income inequality, while a coefficient loser to 0 indicates lower income inequality.

```{r}
#| echo: false
#| message: false
# Calculate the Gini coefficient for Italy's income
italy_gini <- ineq(Italy_income_filtered$thinc_m, type = "Gini")
print(paste("Gini coefficient for Italy:", italy_gini))
```

We can see that for our country of references we find a gini coefficient equal to `r italy_gini` (0.334), which indicates some ineqaulity but not that much. The Gini coefficient is the most widely used measure of income inequality, because it is very easy to calculate and interpret. The problem is that it does not provide any insights into the distribution of income at different parts of the income distribution

### Theil index

We are now going to consider another coefficient to measure inequality: **The Theil Index** The Theil index measures the distance between what we observe in the data and a situation where everyone has the same income. The closer we get to 0 the more equality, the closer to 1, the higher the level of inequality.

```{r}
#| echo: false
#| message: false
# Calculate the Theil index
italy_theil <- Theil(Italy_income_filtered$thinc_m)
print(paste("The Theil index for Italy:", italy_theil))
```

We see a Theil index that again suggests not that high inequality of incomes in Italy. The problem with this index is that it is sensitive to the size of the population and the weights you put in each group, even though in this case we didn't assign any way.

### Palma ratio

A last measure of income inequality is a share ratio, called "Palma ratio", it looks at where most of the inequality comes from, does it come from the fact that the top 10% have a way higher income, and/or from the fact that the bottom 40% have a very low income?

```{r}
#| echo: false
#| message: false
# Calculate the Palma ratio
percentile_90 <- quantile(Italy_income_filtered$thinc_m, 0.9)
percentile_60 <- quantile(Italy_income_filtered$thinc_m, 0.6)
top_10_share <- sum(Italy_income_filtered$thinc_m[Italy_income_filtered$thinc_m >= percentile_90]) / sum(Italy_income_filtered$thinc_m)
bottom_40_share <- sum(Italy_income_filtered$thinc_m[Italy_income_filtered$thinc_m <= percentile_60]) / sum(Italy_income_filtered$thinc_m)
italy_palma <- top_10_share / bottom_40_share

print(paste("The Palma ratio 90/40 for Italy:", italy_palma))
```

This result for Italy, a Palma ratio of approximately 0.65, means that the income share of the top 10% of the population is about 0.65 times larger than the income share of It suggests that a signifiant portion of the total income is concentrated among the wealthiest individuals, something that is not surprising in developed countries like Italy.

The Palma ratio is very useful when looking into extreme income shares, to really understand where most of the inequality comes from. But it has to be used carefully, as it may oversimplify the income distribution by focusing on only two segments of the population (in our case, top 10% and bottom 40%).

## Measuring poverty in Italy

Let's now focus on the bottom of the distribution by looking into different measures of poverty. Even though is not that easy to measure it, especially when having to decide what is the threshold in income to be considered in poverty for example. We can define it in monetary terms by looking at it in **absolute** terms (lack of minimum resources like food), **relative** (to living standards) and **subjective terms** ("do you think you have enough to meet ends?") We first need to find out poverty line, which in Italy changes accordingly to the fact that you live in the north, centre or south and of course it changes accordingly to the number of components in the family and their age. Since we do not know from where the individual come from, how many components in a household, the best thing to do is find a **relative poverty line** calculated as a percentage of the median income. It could be interesting in this case to create a relative poverty line at 60% of the median net income of the household of the sample in the data we are using.

```{r}
#| echo: false
#| message: false
median_income_Italy <- median(Italy_income_filtered$thinc_m)

relative_poverty_line <- median_income_Italy * 0.6

print(paste("Relative Poverty Line:", round(relative_poverty_line, 2)))
```

This poverty line is quite consistent with the poverty line stated by ISTAT in Italy, considered equal to the average monthly pro-capite expenditure in 2017 for a household composed by one component (Instat, 2018).

### Headcount Ratio (poverty rate)

With the poverty line previously determined, we can now look at the poverty rate, calculated as: 
$$
Total'under'the'poverty'line/\\total population
$$

This is what we get:

```{r}
#| echo: false
#| message: false
num_poor <- sum(Italy_income_filtered$thinc_m < relative_poverty_line)

total_population <- nrow(Italy_income_filtered)
poverty_rate <- num_poor / total_population

print(paste("Headcount Ratio (Poverty Rate):", round(poverty_rate * 100, 2), "%"))
```

This is a quite simple and easy to understand measure, but its limitations are obvious. First it does not say anything about the depth of poverty, you are either in poverty or no depending on where the poverty line has been fixed. As we previously said, where the individual are living in Italy changes the poverty line by quite a lot.

### Poverty Gap Index

Let's now have a look into where the poor stand in terms of distance from the poverty line, by using the poverty gap index. This index measures the depth of poverty by considering the distance separating the income of the poor from the poverty line.

It is calculated as:
$$
PGI = {1/N}∑(poverty line - income of poor)/\\poverty line
$$

```{r}
#| echo: false
#| message: false
income_under_poverty_line <- Italy_income_filtered$thinc_m[Italy_income_filtered$thinc_m < relative_poverty_line]
poverty_gap_index <- 1/(length(Italy_income_filtered$thinc_m))* (sum(relative_poverty_line - income_under_poverty_line) / relative_poverty_line)

print(paste("Poverty Gap Index:", round(poverty_gap_index, 1)))
```

It shows that on average, incomes of individuals below the poverty line are 10% below that threshold. Of course with this result only we can't say anything about the inequality inside the population under the poverty line, It might also be a problem the fact that we have many zeros in the income stated by some individuals, which would drag down some calculation and might bias our measures, because those zeros are probably just non replies.

### Poverty in subjective terms
We have measured poverty in relative terms until now, by selecting a threshold to state if a household has to be considered in poverty or no.
Now we are going to look into poverty in more **subjective terms**, by looking into the households ability to make ends meet.
Individuals in the survey have been asked: "Thinking of your household's total monthly income, would you
say that your household is able to make ends meet...". Each of them could answer: 1 (great difficulty), 2 (some difficulty), 3 (fairly easily), 4 (easily).
We are going to consider individuals answering 1 and 2 as in experiencing some level of financial issues, which could be indicative of poverty, and we are going to compute a subjective poverty rate.

```{r}
#| echo: false
#| message: false
poor_households1 <- Italy %>%
  filter(co007_ %in% c(1, 2))

poverty_rate_subjective1 <- nrow(poor_households1) / nrow(Italy)

print(paste("Subjective Poverty Rate:", round(poverty_rate_subjective1 * 100, 2), "%"))
```

We find that 52.58% of individuals in the survey state that their household is able to make ends meet "with great difficulty" or "some difficulty".
Since individuals might overestimate their issue when saying "some difficulty", let's consider in poverty only individuals that make ends meet "with great difficulty":

```{r}
#| echo: false
#| message: false
poor_households <- Italy %>%
  filter(co007_ == 1)

poverty_rate_subjective <- nrow(poor_households) / nrow(Italy)

print(paste("Subjective Poverty Rate:", round(poverty_rate_subjective * 100, 2), "%"))
```

This result is much more consistent with the previous relative poverty measures that we have found so far.

## Measuring inequality based on health outcomes

Here we are going to measure health inequality, but let's first have a look at the self perceived health, where respondents had to answer about their health status with: 1) Excellent, 2) Very good, 3) Good, 4) Fair, 5) Poor

```{r}
#| echo: false
#| message: false
Italy_satisfaction_filtered <- Italy |>
  filter(sphus>0)
health_satisfaction_freq <- table(Italy_satisfaction_filtered$sphus)

# Compute percentages
health_satisfaction_percent <- prop.table(health_satisfaction_freq) * 100

health_satisfaction_table <- data.frame(
  Satisfaction_Level = as.character(names(health_satisfaction_freq)),
  Frequency = as.numeric(health_satisfaction_freq),
  Percentage = as.numeric(health_satisfaction_percent)
)

# Sort the table by Satisfaction_Level
health_satisfaction_table <- health_satisfaction_table[order(health_satisfaction_table$Satisfaction_Level), ]

# Print the table
print(health_satisfaction_table)
```
We can see that most of the respondents answer between fair and good when asked about their self reported health, while most of them restrain from answering "excellent" to that question.

Before getting into the measure of health inequality, let's look into a variable that is less subjective than the previous one. chronic_mod is a variable that looks into the number of chronic disease of individuals. Respondents had to answer to the question: "Has a doctor ever told you that you had …", considering the following chronic diseases: 1. A heart attack; 2. High blood pressure or hypertension; 3. High blood cholesterol; 4. A stroke or cerebral vascular disease; 5. Diabetes or high blood sugar; 6. Chronic lung disease (...); 10. Cancer or malignant tumor (…); 11. Stomach or duodenal ulcer, peptic ulcer; 12. Parkinson disease; 13. Cataracts; 14. Hip fracture or femoral fracture.
Individuals that did not have any of those have 0 as result, so we can look into the percentage of respondents that has had a chronic disease and that has not:


```{r}
#| echo: false
#| message: false
with_chronic <- sum(Italy$chronic_mod > 0)
total_responses <- nrow(Italy)
percentage_with_chronic <- with_chronic / total_responses * 100

# Calculate the percentage of respondents with no chronic disease
without_chronic <- sum(Italy$chronic_mod == 0)
percentage_without_chronic <- without_chronic / total_responses * 100

# Create a data frame for plotting
chronic_status <- data.frame(
  Status = c("With Chronic Disease", "Without Chronic Disease"),
  Percentage = c(percentage_with_chronic, percentage_without_chronic)
)

# Plot the results
ggplot(chronic_status, aes(x = Status, y = Percentage, fill = Status)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of Respondents with Chronic Diseases") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("With Chronic Disease" = "red", "Without Chronic Disease" = "lightgreen"))
```

The data shows that more than 60% of respondents for Italy have had at least one of the chronic disease mentioned, it looks quite consistent with the low level of self reported health we saw before.

With this general view of our country of reference, we are now going measure health inequality at the individual level.

### Measures of health inequality
We first have to select some health indicators, from the data available we can consider:
-   *Self-perceived health status*
-   *Presence of chronic diseases*
-   *Limitations in activities of daily living (ADL)*, which indicates difficulties in performing basic self-care tasks, from 0 to 5, the higher the index the higher the difficulty of the respondent to complete these tasks.
-   *Limitations in instrumental activities of daily living (IADL)*, which indicates the difficulty of performing more difficult activities necessary for independent living, ranges from 0 to 5, the higher the more difficulties in doing these tasks.
-   *Depression scale*, which measures the current depression, based on several measures: depressed mood, pessimism, suicidality, guilt, sleep, interest, irritability, appetite, fatigue, concentration, enjoyment and tearfulness. It ranges from 0 (not depressed), to 12 (very depressed)

Then we will have to find some subgroups, in our case what would be interesting to see is in terms of education, sex and place of residence (City, village, ...)

From here what we want to do is create a Health Utility Index, as done by Asada et al. in 2014. To create this we have to normalise hour health indicators into a common scale (range from 0 to 1), as the Health Utility Index assigns being dead a value of zero and full health a value of one (Asada et al, 2014).

```{r}
#| echo: false
#| message: false
self_reported_scaled <- ifelse(Italy$sphus >= 0, (Italy$sphus - 1) / 4, NA)
chronic_disease_scaled <- ifelse(Italy$chronic_mod >= 0, ifelse(Italy$chronic_mod > 0, 1, 0), NA)
ADL_scaled <- ifelse(Italy$adla >= 0, Italy$adla / 5, NA)
IADL_scaled <- ifelse(Italy$iadlza >= 0, Italy$iadlza / 5, NA)
depression_scaled <- ifelse(Italy$eurod >= 0, Italy$eurod / 12, NA)

Health_Utility_Index <- rowMeans(cbind(self_reported_scaled, chronic_disease_scaled, ADL_scaled, IADL_scaled, depression_scaled), na.rm = TRUE)

HUI <- Health_Utility_Index

Italy$HUI <- HUI
```

Having the HUI for each respondent we can now look into different subgroups in our data, to grasp health inequalities between them.

Let's start with gender:
```{r warning=FALSE}
#| echo: false
#| message: false
ggplot(data = Italy, aes(x = Gender, y = HUI, fill = Gender)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Health Utility Index (HUI)", title = "Distribution of HUI by Gender") +
  scale_fill_manual(values = c("#FF9999", "#66CCFF")) +
  theme_minimal()
```
We don't see any meaningful difference in the distribution between female and male, considering the HUI that we created.

```{r}
education_labels <- c("None", "Primary", "Lower Secondary", "Secondary", "Post-Secondary", "First Stage Tertiary", "Second Stage Tertiary", "Still in School", "Other")

# Convert education levels to factor with corresponding labels
Italy$isced1997_r <- factor(Italy$isced1997_r, levels = c(0:6, 95, 96), labels = education_labels)

average_hui <- Italy %>%
  group_by(isced1997_r) %>%
  summarise(avg_HUI = mean(HUI, na.rm = TRUE))

ggplot(data = average_hui, aes(x = isced1997_r, y = avg_HUI)) +
  geom_bar(stat = "identity", fill = "grey") +
  labs(x = "Education Level", y = "Average Health Utility Index (HUI)", title = "Average HUI by Education Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
Italy$iv009_mod <- replace(Italy$iv009_mod, Italy$iv009_mod < 0, NA)
location_labels <- c("Big City", "Suburbs or Outskirts of a Big City", "Large Town", "Small Town", "Rural Area or Village")

# Convert area of location to factor with corresponding labels
Italy$iv009_mod <- factor(Italy$iv009_mod, levels = 1:5, labels = location_labels)

# Calculate average HUI for each area of location
average_hui_location <- Italy %>%
  group_by(iv009_mod) %>%
  summarise(avg_HUI = mean(HUI, na.rm = TRUE))

# Create a bar graph showing average HUI for each area of location
ggplot(data = average_hui_location, aes(x = iv009_mod, y = avg_HUI)) +
  geom_bar(stat = "identity", fill = "#FF9999") +
  labs(x = "Area of Location", y = "Average Health Utility Index (HUI)", title = "Average HUI by Area of Location") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

