---
title: "SHARE_Italy"
format:
  pdf:
    toc: true
    toc-depth: 2
    toc-title: Contents
    number-sections: true
    colorlinks: true
editor: visual
author: Sula Disem
---

```{r}
#| echo: false
#| message: false
here::i_am("Final_SHARE_Italy.Rproj")
library(here)
library(ggplot2)
library(tidyr)
library(dplyr)
library(vroom)
library(stringr)
library(knitr)
library(patchwork)
library(broom)
```

### Link to the github repository of this project {.appendix}

[Github repository](https://github.com/suladi23/measurement_issues_SHARE_Italy)

### Source of the data {.appendix}
This paper uses data from the generated easySHARE data set (DOI: [10.6103/SHARE.easy.800](https://share-eric.eu/data/data-set-details/easyshare)), see Gruber et al. (2014) for methodological details. The easySHARE release 8.0.0 is based on SHARE Waves 1, 2, 3 (SHARELIFE), 4, 5, 6, 7 and 8 (DOIs: [10.6103/SHARE.w1.800](http://dx.doi.org/10.6103/SHARE.w1.800), [10.6103/SHARE.w2.800](http://dx.doi.org/10.6103/SHARE.w2.800), [10.6103/SHARE.w3.800](http://dx.doi.org/10.6103/SHARE.w3.800), [10.6103/SHARE.w4.800](http://dx.doi.org/10.6103/SHARE.w4.800),[10.6103/SHARE.w5.800](http://dx.doi.org/10.6103/SHARE.w5.800), [10.6103/SHARE.w6.800](http://dx.doi.org/10.6103/SHARE.w6.800), [10.6103/SHARE.w7.800](http://dx.doi.org/10.6103/SHARE.w7.800), [10.6103/SHARE.w8.800](http://dx.doi.org/10.6103/SHARE.w8.800))

### References {.appendix}
Börsch-Supan, A. & S. Gruber (2020): easySHARE. Release version: 8.0.0. SHARE-ERIC. Dataset. doi: [10.6103/SHARE.easy.800](http://dx.doi.org/10.6103/SHARE.easy.800)


```{r}
#| echo: false
#| message: false
load("easySHARE_rel8_0_0.rda")
```

```{r}
#| echo: false
#| message: false
wave7 <- easySHARE_rel8_0_0 |>
  filter(wave == "7")
```

```{r}
#| echo: false
#| message: false
Italy <- wave7 |>
  filter(country == "16")
```

```{r}
#| echo: false
#| message: false
France <- wave7 |>
  filter(country == "17")
```

```{r}
#| echo: false
#| message: false
Germany <- wave7 |>
  filter(country == "12")
```


## Description of the Dataset for Italy

The SHARE survey, or Survey of Health, Ageing and Retirement in Europe, is a multidisciplinary and cross-national research initiative aimed at understanding the dynamics of health, aging, and retirement among older adults in Europe and beyond.
It aims to provide insights into the health, social, economic, and environmental factors influencing individuals' well-being and quality of life as they age.
To do so it collects data from the same individual over multiple waves at regular intervals, to track changes in various aspect of life over time. The data is collected from a wide range of European countries, to enable comparison with trends and patterns across different cultures and economic contexts.

SHARE employs a rigorous **sampling design** to ensure representativeness of the older population in each participating country. The survey includes individuals from diverse socio-economic backgrounds and geographic regions.

```{r warning=FALSE}
#| echo: false
#| message: false
nrows_Italy=nrow(Italy)
ncols_Italy=ncol(Italy)
```

After some manipulation of the data, filtering it to use only "Wave 7" and to consider only the country of choice "Italy", the Dataset we will be using has **`r nrows_Italy` rows** and **`r ncols_Italy` columns (variables)**


```{r}
#| echo: false
#| message: false
#Mainupulating the data to be able to work better afterwards

Italy$ep005_[Italy$ep005_ == 1] <- "Retired"
Italy$ep005_[Italy$ep005_ == 2] <- "Employed or Self-employed"
Italy$ep005_[Italy$ep005_ == 3] <- "Unemployed"
Italy$ep005_[Italy$ep005_ == 4] <- "Permanently Sick or Disabled"
Italy$ep005_[Italy$ep005_ == 5] <- "Homemaker"
Italy$ep005_[Italy$ep005_ == 97] <- "Other"

Italy$female[Italy$female == 0] <- "Male"
Italy$female[Italy$female == 1] <- "Female"

Italy <- Italy %>%
  rename(
    Gender = female,
    `Employment status` = ep005_
  )
```


The following table provides some descriptive statistics of Income, Age and Years of education.

```{r}
#| echo: false
#| message: false
# Calculate descriptive statistics for household income
Italy_income_filtered <- Italy %>% 
  filter(thinc_m != -10)
summary_income <- summary(Italy_income_filtered$thinc_m)

# Calculate descriptive statistics for gender
gender_stats <- table(Italy$Gender)

# Calculate descriptive statistics for age
summary_age <- summary(Italy$age)

# Calculate descriptive statistics for years of education
Italy_education_filtered <- Italy |>
  filter(eduyears_mod >= 0)
summary_education <- summary(Italy_education_filtered$eduyears_mod)

# Calculate descriptive statistics for employment status
Italy_employment_filtered <- Italy |>
  filter(`Employment status` != "-12" & `Employment status` != "-15")
employment_status <- table(Italy_employment_filtered$`Employment status`)
```

```{r}
#| echo: false
#| message: false
summary_combined <- data.frame(
  Variable = c("Income", "Age", "Years of education"),
  Min = c(summary_income["Min."], summary_age["Min."], summary_education["Min."]),
  `1st Qu.` = c(summary_income["1st Qu."], summary_age["1st Qu."], summary_education["1st Qu."]),
  Median = c(summary_income["Median"], summary_age["Median"], summary_education["Median"]),
  Mean = c(summary_income["Mean"], summary_age["Mean"], summary_education["Mean"]),
  `3rd Qu.` = c(summary_income["3rd Qu."], summary_age["3rd Qu."], summary_education["3rd Qu."]),
  Max = c(summary_income["Max."], summary_age["Max."], summary_education["Max."])
)
kable(summary_combined, format = "markdown")
```

Out of `r nrows_Italy` observations, we have here a table showing how the individuals are distributed among different types of employment status and gender:
```{r}
#| echo: false
#| message: false
combined <- data.frame(
  Count = c(employment_status, gender_stats))
kable(combined, format = "markdown")
```


