---
title: "SHARE_Italy"
format: pdf
editor: visual
author: Sula Disem
---

```{r}
#| echo: false
#| message: false
here::i_am("Final_SHARE_Italy.Rproj")
library(here)
library(ggplot2)
library(tidyr)
library(dplyr)
library(vroom)
library(stringr)
library(knitr)
library(patchwork)
library(broom)
```

### Link to the github repository of this project {.appendix}

[Github repository](https://github.com/suladi23/measurement_issues_SHARE_Italy)

### Source of the data {.appendix}
This paper uses data from the generated easySHARE data set (DOI: [10.6103/SHARE.easy.800](https://share-eric.eu/data/data-set-details/easyshare)), see Gruber et al. (2014) for methodological details. The easySHARE release 8.0.0 is based on SHARE Waves 1, 2, 3 (SHARELIFE), 4, 5, 6, 7 and 8 (DOIs: [10.6103/SHARE.w1.800](http://dx.doi.org/10.6103/SHARE.w1.800), [10.6103/SHARE.w2.800](http://dx.doi.org/10.6103/SHARE.w2.800), [10.6103/SHARE.w3.800](http://dx.doi.org/10.6103/SHARE.w3.800), [10.6103/SHARE.w4.800](http://dx.doi.org/10.6103/SHARE.w4.800),[10.6103/SHARE.w5.800](http://dx.doi.org/10.6103/SHARE.w5.800), [10.6103/SHARE.w6.800](http://dx.doi.org/10.6103/SHARE.w6.800), [10.6103/SHARE.w7.800](http://dx.doi.org/10.6103/SHARE.w7.800), [10.6103/SHARE.w8.800](http://dx.doi.org/10.6103/SHARE.w8.800))

### References {.appendix}
Börsch-Supan, A. & S. Gruber (2020): easySHARE. Release version: 8.0.0. SHARE-ERIC. Dataset. doi: [10.6103/SHARE.easy.800](http://dx.doi.org/10.6103/SHARE.easy.800)


```{r}
#| echo: false
#| message: false
load("easySHARE_rel8_0_0.rda")
```

```{r}
#| echo: false
#| message: false
wave7 <- easySHARE_rel8_0_0 |>
  filter(wave == "7")
```

```{r}
#| echo: false
#| message: false
Italy <- wave7 |>
  filter(country == "16")
```

```{r}
#| echo: false
#| message: false
France <- wave7 |>
  filter(country == "17")
```

```{r}
#| echo: false
#| message: false
Netherlands <- wave7 |>
  filter(country == "14")
```
